### 2004-05-19 Documentation/part2.tex
*** empty log message ***

{\tt \$numofbins = 20}\\

@@

% WEITER:

Based on the definition of channels it is possible to assemble the 
signal and background event numbers for each bin.  
The signal $s_i$ in one bin for a rule now can be composed out of 
several channels by (OFTEN ONLY ONE) 
\begin{equation}
s_i=\epsilon_{c_1}\cdot n_i^{c_1}\,+\,\epsilon_{c_2}\cdot n_i^{c_2}\,+\,\ldots
\end{equation}
where the $\epsilon$'s are weight factors determined by the properties
of the detector.
Similarly the background in one bin $i$ can be specified by
\begin{equation}
b_i=\epsilon_{c_1}\cdot n_i^{c_1}\,+\,\epsilon_{c_2}\cdot n_i^{c_2}\,+\,\ldots
\end{equation}
This basic building blocks of a rule are specified by
\begin{quote}
{\tt \tb @signal = 0.5 @ \#channel\_1\\
\tb @background = 0.001 @ \#channel\_2 :  0.005 @ \#channel\_3
}
\end{quote}

Quite often it is not sufficient to have constant efficiencies and therefore
it is possible to define efficiencies for each bin and channel $\alpha_i^c$
and thus the event rates in one bin become
\begin{equation}
s_i=\epsilon_{c_1}\alpha_i^{c_1}\cdot n_i^{c_1}\,+
\,\epsilon_{c_2}\alpha_i^{c_2}\cdot n_i^{c_2}\,+\,\ldots\,,
\end{equation}
and analogous for $b_i$.

The set of all $s_i$'s and $b_i$'s now can be composed in a form that takes
some types of systematical error into account. The two most important and
most easily parameterized systematics are a normalization and an energy
calibration error. These error are considered separately for the signal events
and the background events. The implementation of the normalization error
is straight forward
\begin{equation}
s_i(a):=a\cdot s_i
\end{equation} 
with an analogous definition for the background events.

For the parameterization of an energy calibration error two possibilities
are implemented. The first one being somewhat simpler, whereas the second one
is more accurate but requires a careful choice of parameters. The first 
option is
\begin{equation}
s_i(a,b):=s_i(a)+b\cdot s_i\, E_i/(E_\mathrm{max}-E_\mathrm{min})
\end{equation}
and it is refered to as tilt since it describes a linear distortion, a tilt
of the event rate spectrum. The second option is closer to an actual energy
calibration error, which basically amounts to replacing the reconstructed 
energy $E'$ by $(1+b)\cdot E'$ and we use following approximation
\begin{eqnarray}
s_i(b)&=& (1+b)\cdot(s_{k+1}-s_k)\cdot\delta+s_i\,,\\
\delta&=&b\cdot(i+t_0+ 1/2)+i\,,\nonumber\\
k&=&\mod(\delta,1)\,,\nonumber\\
t_0&=&E_\mathrm{min}/\Delta E_0\,.\nonumber
\end{eqnarray}
Special care is required at the boundaries when $k<1$ or $k+1>N_\mathrm{bins}$,
since there $s_k$ or $s_{k+1}$ may not have been calculated. One possibility
to deal with those cases is to assume that for those values of $k$ $s_k$ is
zero. This is the default. In cases where $s$ at the boundary is however still
sizeable this may lead to intolerable errors and subsequently to a wrong
estimate of the impact of a calibration error. Therefore in those cases it
is advisable to truncate the analysis range by a few bins at the boundary
and therefore ensure in this way that only those $s_i$ are used whose index
$k$ is within the range $1,\ldots, N_\mathrm{bins}-1$.

In order to perform this fine tuning, it is possible to constrain 
the energy range from which the data are taken for the 
analysis\index{Energy window}. This is achieved by setting the energy window
variable. 
\begin{quote}
{\tt 
\tb @energy\_window = 4.0 : 50.0 
}
\end{quote}
The purpose is mainly to allow a fine tuning of the description
of an energy calibration error. The default setting is that the energy window
coincides with the range defined by the minimal and maximal reconstructed 
energy. For the error dimension using the calibration method (C) it is 
advisable to reduce the energy window by two or three bins at both the lower
and upper end. As a rule of
thumb consider the three standard deviations error on the calibration and
evaluate the corresponding shift in $E'$ at the lower and upper energy limit 
and reduce the analysis range by this amount.

Thus the total event rate $x_i$ in a bin $i$ is given by
\begin{equation}
x_i(a,b,c,d)=s_i(a,b)+b_i(c,d)
\end{equation}
and thus a function of four parameters. The central values for all of
the four parameters are always needed. They are called signal normalization
($a$), signal tilt/calibration ($b$), background  normalization ($c$) and
background tilt/calibration ($d$). The default values are
\begin{equation}
a=1\,,\quad b=0\,,\quad c={\tt NaN}\,,\quad d=0\,,
\end{equation}
thus for the background normalization $c$ a value has to be specified in 
\emph{all} cases. The central values for the normalization and the 
corresponding central value of tilt/calibration are regarded
as a pair. The errors are treated in the same way.
\begin{quote}
{\tt
\tb @signalerror =       0.001  :       0.01\\
\tb @backgroundcenter =  0.1 :       0.0\\
\tb @backgrounderror =   0.001 :       0.01
}
\end{quote}
There is no {\tt @signalcenter} since by default the central value for the
signal normalization is $1$ and the central value for the tilt/calibration 
is $0$.

The four parameters $a,b,c,d$ have been introduced in order to describe
systematical uncertainties, they are so called nuisance parameters.
For the analysis of the systematical errors the so called 
 pull method is 
used~\cite{Fogli:2002pt}\footnote{In fact the pull method
was employed already in~\cite{Huber:2002mx} before 
\cite{Fogli:2002pt} appeared.}. 
In the  pull method $k$ systematical errors are included by introducing 
$k$ additional variables $\zeta_k$, which will  be called 
nuisance parameters in the following. 
The nuisance parameters describe the dependence of the event rates on the 
various sources of systematical errors, \eg\ an error on the total 
normalization is included by multiplying the expected number of events in 
each bin by a factor $(1+\zeta_1)$. The variation of $\zeta_1$ in the fit 
is constrained by adding a penalty $p_1$ to the $\chi^2$-function. In case 
of a  Gau\ss ian distributed systematical error this penalty is 
given by
\begin{equation}
\label{eq:penalty}
p_i=\frac{(\zeta_i-\zeta_i^0)^2}{\sigma_{\zeta_i}^2}\,,
\end{equation}
where $\zeta_i^0$ denotes the mean and $\sigma_{\zeta_i}$ the standard 
deviation of the corresponding nuisance parameter, \ie\ the amount of 
systematical uncertainty. 
The resulting $\chi^2$ is then minimized with respect to all nuisance 
parameters $\zeta_i$ and this yields $\chi^2_\mathrm{pull}$
\begin{equation}
\chi^2_\mathrm{pull}(\boldsymbol{\lambda}):=\min_{\{\zeta_i\} } \,\, \left( 
\chi^2(\boldsymbol{\lambda},
\zeta_1, \ldots, \zeta_k)+ \sum_{j=1}^{k} p_j(\zeta_j)\right)\,,
\end{equation}
where $\boldsymbol{\lambda}$ denotes the oscillation parameters 
including the matter density
$\rho$. One advantage of the pull method is that whenever the number $N$ of 
data points is much larger than $k$, it is numerically easier to compute 
$\chi^2_\mathrm{pull}$ than to invert the $N\times N$ covariance matrix. For
the experiments considered here $N$ is typically $20$ and $k\sim 4$, thus
the pull method is numerically much faster. Moreover it is more flexible and 
allows the inclusion of systematical errors also for a 
Poissonian $\chi^2$-function.
In~\cite{Fogli:2002pt} it was shown that the pull method and the covariance
based approach are equivalent for a Gau\ss ian and linear model. In general
there is a separate $(\chi^2_\mathrm{pull})^\alpha$ for each rule $\alpha$, 
\ie\ pair of signal and background spectra, with a separate set of 
nuisance parameters $\zeta_i^\alpha$. In this case $\chi^2_\mathrm{pull}$
is the sum of all individual  $(\chi^2_\mathrm{pull})^\alpha$.
In this way the dependence on the $k$ nuisance parameters has been eliminated 
from $\chi^2_\mathrm{pull}$.

The user has the possibility to choose the set $\{\zeta_i\}$ of nuisance 
parameters which are marginalized. This achieved with the error dimension 
variable\index{Error dimension}, and the different possibilities are shown in
table~\ref{tab:error_dim}. If a parameter is designated with $+$, it will
be marginalized and therefore the corresponding error needs to have a non-zero
value. In the cases labeled (number 4 and 8) ``total rates'' 
the summation over the bins is performed \emph{before} computing 
the $\chi^2$. The case labeled (number 7) ``spectrum only'' leaves the 
normalization free ($\sigma_a=\sigma_c=\infty$) and therefore counts only the 
spectral information in the data. As a consequence
the settings for the normalization error will be ignored. It is furthermore
possible to extend the set of error dimensions and thus the set of possible 
systematical errors to be studied, but this requires proper C programing, 
compiling and linking with \GLOBES. The interface for user defined error 
dimensions is described in detail in section~\ref{sec:ud_error_dim}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{center}
\begin{table}[hbt!]
\begin{center}
\begin{tabular}[h]{|c|cccc|c|c|}
\hline
Error dimension&$a$&$b$&$c$&$d$&Tilt/Calibration&Remarks\\
\hline
\hline
0&+&+&+&+&T&\\
2&-&-&-&-&-&\\
4&+&+&+&+&T&total rates\\
7&o&-&o&-&-&spectrum only\\
8&-&-&-&-&-&total rates\\
9&+&+&+&+&C&\\
\hline
\end{tabular}
\caption[Table of error dimensions]{\label{tab:error_dim}
Values of the error dimension variable and their meaning.
 }
\end{center} 
\end{table} 
\end{center}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The error dimensions can be chosen for each rule by
\begin{quote}
{\tt 
\tb @errordim = 1
}
\end{quote}

Next we can assemble the various pieces into one rule
\begin{quote}
{\tt rule(\#rule\_1)<\\
\tb @signal = 0.5 @ \#channel\_1\\
\tb @background = 0.001 @ \#channel\_2 :  0.005 @ \#channel\_3\\
\tb @signalerror =       0.001  :       0.01\\
\tb @backgroundcenter =  0.1 :       0.0\\
\tb @backgrounderror =   0.001 :       0.01\\
\tb @errordim = 1\\
\tb @energy\_window = 4.0 : 50.0\\ 
>}
\end{quote}

>>>>

{\tt \$bins = 20}\\

@@

For each rule, the signal event rate $s_i$ in the $i$th bin can be composed 
out of one or more channels by 
\begin{equation}
s_i=\alpha_{c_{s1}}\cdot n_i^{c_{s1}}\,+\,\alpha_{c_{s2}}\cdot n_i^{c_{s2}}\,+\,\ldots
\end{equation}
where the $\alpha$'s are overall normalization factors/efficiencies
 determined by the properties of the detector. Note that bin-based (energy-dependent) efficiencies can be defined with the post-smearing efficiencies in the last section. 
In addition, note that in most cases, it makes sense to have only one 
signal channel and to assign all sorts of perturbations to the background. 
%
Similarly, the background event rate $b_i$ in the $i$th bin can be composed
out of one or more channels
\begin{equation}
b_i=\beta_{c_{b1}}\cdot n_i^{c_{b1}}\,+\,\beta_{c_{b2}}\cdot n_i^{c_{b2}}\,+\,\ldots \, ,
\end{equation}
where the channels can be any combination of the ones in the signal rate and 
additional ones. The background normalization factors very often have
a specific meaning. For example, they may correspond to a fraction
of mis-identified events (charge or flavor mis-identification).
%
These basic building blocks of each rule are, within the rule environment,
for example defined by
\begin{quote}
{\tt \tb @signal = 0.5 @ \#channel\_1\\
\tb @background = 0.001 @ \#channel\_2 :  0.005 @ \#channel\_3
}
\end{quote}

For the analysis of the systematical errors, the so called 
 ``pull method'' is used~\cite{Fogli:2002pt}\footnote{In fact the pull 
method was employed already in \Ref~\cite{Huber:2002mx} before 
\Ref~\cite{Fogli:2002pt} appeared.}. 
For the pull method, $k$ systematical errors are included by introducing 
$k$ additional variables $\zeta_k$, which are the 
so-called ``nuisance parameters''. 
The nuisance parameters describe the dependence 
of the event rates on the various sources of systematical errors, 
such as an error on the total 
normalization is included by multiplying the expected number of events in 
each bin by a factor $(1+\zeta_1)$. The variation of $\zeta_1$ is in the fit 
constrained by adding a penalty $p_1$ to the $\chi^2$-function. In case 
of a  Gau\ss ian distributed systematical error, this penalty is 
given by
\begin{equation}
\label{eq:penalty}
p_i=\frac{(\zeta_i-\zeta_i^0)^2}{\sigma_{\zeta_i}^2}\,,
\end{equation}
where $\zeta_i^0$ denotes the mean and $\sigma_{\zeta_i}$ the standard 
deviation of the corresponding nuisance parameter. We further on also refer
to the mean as the ``central value'', and to the standard deviation as
the ``error''. The latter corresponds to the actual systematical uncertainty. 
The resulting $\chi^2$ is then minimized with respect to all nuisance 
parameters $\zeta_i$, which leads to $\chi^2_\mathrm{pull}$
\begin{equation}
\chi^2_\mathrm{pull}(\boldsymbol{\lambda}):=\min_{\{\zeta_i\} } \,\, \left( 
\chi^2(\boldsymbol{\lambda},
\zeta_1, \ldots, \zeta_k)+ \sum_{j=1}^{k} p_j(\zeta_j)\right)\,.
\end{equation}
Here $\boldsymbol{\lambda}$ refers to the oscillation parameters 
including the matter density
$\rho$. One advantage of the pull method is that whenever the number $N$ of 
data points is much larger than $k$, it is numerically easier to compute 
$\chi^2_\mathrm{pull}$ than to invert the $N\times N$ covariance matrix. For
the experiments considered here, $N$ is typically $20$ and $k\sim 4$, 
which means that the pull method is numerically much faster. Moreover,
 it is more flexible and  allows the inclusion of systematical errors 
also for a  Poissonian $\chi^2$-function.
In~\cite{Fogli:2002pt}, it has been demonstrated that the pull method 
and the covariance based approach are equivalent for a Gau\ss ian and 
linear model. In general,
there is a separate $(\chi^2_\mathrm{pull})^r$ for each rule $r$, 
\ie , pair of signal and background spectra, with a separate set of 
nuisance parameters $\zeta_i^\alpha$. Thus, $\chi^2_\mathrm{pull}$
is the sum of all individual  $(\chi^2_\mathrm{pull})^r$'s.
By the minimization, the dependence on the $k$ nuisance parameters has been eliminated  from $\chi^2_\mathrm{pull}$.

Now we can introduce the different systematical errors. 
The two most important and
most easily parameterized systematical errors are the normalization 
and an energy calibration errors. These errors are assumed to be independent between the signal events and the background events, which means that
this systematics treatment defines the grouping into signal or background.
The implementation of the normalization error
is straightforward:
\begin{equation}
s_i(a):=a\cdot s_i
\end{equation} 
with an analogous definition for the background events. Here, $a$ is the``nuisance'' parameter, which will be minimized over later.

For the parameterization of an energy calibration error, two possibilities
are implemented. The first one (method ``T'') is somewhat simpler, 
whereas the second one (method ``C'')
is more accurate, but it requires a careful choice of parameters. 
The first option (method ``T'') is
\begin{equation}
s_i(a,b) \equiv s_i(a)+b\cdot s_i\, E'_i/(E'_\mathrm{max}-E'_\mathrm{min}),
\end{equation}
where $E'_{\mathrm{min}}$ and $E'_{\mathrm{max}}$ correspond to {\tt \$emin}
and {\tt \$emax}, and $E'_i$ is the mean (reconstructed) 
energy of the $i$th bin. It is often refered to as a ``tilt'' of the
spectrum, since it describes a linear distortion 
of the event rate spectrum. Note that this definition of a tilt makes
only sense in combination with a large enough normalization error, since
the tilt also affects the normalization.
%
The second option (method ``C'') is closer to an actual energy
calibration error, which means that one should test this option whenever
one suspects a large impact of this systematical error.
It is based upon replacing the events in the $i$th bin by the ones at
the energy $(1+b)\cdot E'_i$. If this target energy does not exactly hit
a (discrete) bin energy $E_k$, linear interpolation is used. We use the following approximation:
\begin{eqnarray}
s_i(a,b)&=& (1+b)\cdot \left[ \left( s_{k+1}(a)-s_k(a) \right)\cdot (\delta-k) +s_k(a) \right] \,,\\
\delta&=&b\cdot(i+t_0+ 1/2)+i\,,\nonumber\\
k&=& \mathrm{div}(\delta,1)\,,\nonumber\\
t_0&=&E'_\mathrm{min}/\Delta E_0\,.\nonumber
\end{eqnarray}
Here $\Delta E_0$ is the bin width ({\tt \$emax}-{\tt \$emin})/{\tt \$bins}
and ``div'' refers to the integer part of the division. Note that the
factor $(1+b)$ in the first equation comes from a renormalization of
the bin width, since also the bin width is altered by the replacement of
the energies. Furthermore, special care has to be payed to the limits
 $k<1$ or $k+1>N_\mathrm{bins}$, since there $s_k$ or $s_{k+1}$ may not have been calculated. By default, it is assumed that $s_k$ is
zero in those cases. However, if the event rates are still large at
the limits, errors will be introduced leading to a wrong
estimate of the impact of the calibration error. In this case,
one should truncate the analysis range by a few bins at the boundaries
and therefore ensure in this way that only those $s_i$ are used whose index
$k$ is within the range $0,\ldots, N_\mathrm{bins}-1$ (\cf, \figu{energies}). 
Thus, it is possible to constrain the analysis energy range 
with each rule to an energy window\index{Energy window}:
\begin{quote}
{\tt 
\tb @energy\_window = 4.0 : 50.0 
}
\end{quote}
The default energy window is given between the minimal and maximal reconstructed energy. To be on the save side, reduce analysis window
compared to the bin range on each side by about three standard deviations 
of the calibration error.

Eventually, the total event rate $x_i$ in a bin $i$ is given by
\begin{equation}
x_i(a,b,c,d)=s_i(a,b)+b_i(c,d) \, ,
\end{equation}
and is thus a function of four parameters. 
The four parameters $a,b,c,d$ have been introduced in order to describe
systematical uncertainties and are the nuisance parameters.
Each of the four parameters has a central value and systematical error.
The central values for all of
the four parameters have to be always defined. They are called 
signal normalization ($a$), signal tilt/calibration ($b$), 
background  normalization ($c$) and
background tilt/calibration ($d$). The default values are
\begin{equation}
a=1\,,\quad b=0\,,\quad c={\tt not \, \, assigned}\,,\quad d=0\,.
\end{equation}
Thus, for the background normalization $c$, the value has to be specified 
in \emph{either} case. The values for the normalization and the 
values of tilt/calibration are always regarded
as a pairs, \ie, they are given in the form {\tt normalization : tilt}. The errors are treated in the same way. For example, we have
\begin{quote}
{\tt
\tb @signalerror =       0.001  :       0.01\\
\tb @backgroundcenter =  0.1 :       0.0\\
\tb @backgrounderror =   0.001 :       0.01
}
\end{quote}
There is no {\tt @signalcenter} in this definition, 
since by default the central value for the
signal normalization is $1$ and the central value for the tilt/calibration 
is $0$.

The user has the possibility to choose the set $\{\zeta_i\}$ of nuisance 
parameters which are minimized over. This choice is specified with the 
error dimension variable\index{Error dimension}, and the different possibilities are shown in
\Tab~\ref{tab:error_dim}. The error dimensions can be chosen for each rule 
by {\tt @errordim}. For example, 
\begin{quote}
{\tt 
\tb @errordim = 1
}
\end{quote}
In general, it is possible to extend the set of error dimensions or the 
set of possible systematical errors, but this requires C programming, 
compiling and linking with \GLOBES. 
%The interface for user defined error 
%dimensions is described in detail in section~\ref{sec:ud_error_dim}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{center}
\begin{table}[t!]
\begin{center}
\begin{tabular}[h]{|c|cccc|c|l|}
\hline
Error dimension&$a$&$b$&$c$&$d$&Tilt/Calibration&Remarks\\
\hline
\hline
0&+&+&+&+&T&Systematics with tilt\\
2&-&-&-&-&-&No systematics\\
4&+&+&+&+&T&Total rates\\
7& $\infty$ &-& $\infty$ &-&-&Spectrum only\\
8&-&-&-&-&-&Total rates, No systematics\\
9&+&+&+&+&C&Systematics with calibration\\
\hline
\end{tabular}
\caption[Table of error dimensions]{\label{tab:error_dim}
Possible values of the error dimensions variable in \GLOBES\ and their meaning. If a parameter is designated with $+$, it will
be marginalized over, and therefore the corresponding error needs to 
have a non-zero value. If the cases with ``total rates'' in the remarks, 
the summation over the bins is performed \emph{before} computing 
the $\chi^2$, \ie, no spectral information is used. 
The error dimension~7 ``spectrum only'' leaves the 
normalization free ($\sigma_a=\sigma_c=\infty$), and therefore 
only the spectral information is used. As a consequence,
the settings for the normalization error will be ignored (designated 
with the symbol $\infty$). 
 }
\end{center} 
\end{table} 
\end{center}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Eventually, a rule looks like
\begin{quote}
{\tt rule(\#rule\_1)<\\
\tb @signal = 0.5 @ \#channel\_1\\
\tb @background = 0.001 @ \#channel\_2 :  0.005 @ \#channel\_3\\
\tb @signalerror =       0.001  :       0.01\\
\tb @backgroundcenter =  0.1 :       0.0\\
\tb @backgrounderror =   0.001 :       0.01\\
\tb @errordim = 1\\
\tb @energy\_window = 4.0 : 50.0\\ 
>}
\end{quote}
